Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2025-02-19 demystify.tex
updates

International students, especially those from less well-known universities and smaller countries, often wonder if they should apply to a CS PhD program in the US. Common concerns include (i) the \textbf{difficulty of getting admitted}, (ii) the \textbf{cost of graduate study} in the US, and for some, (iii) the fear of \textbf{discrimination} e.g., based on gender identity or cultural background.

@@

\paragraph{Discrimination is not tolerated} Not only is discrimination illegal, diversity and inclusion are highly valued in US universities. Many universities take great pride in their diverse student body and faculty.

In fact, many resources are available to encourage and support students from minority and diverse groups to pursue higher education and research, e.g., specific scholarships and fellowships. Faculty in CS are also familiar with and are encouraged to support such students in their group. There are also incentives, such as dedicated funding and awards, to motivate faculty to recruit and mentor minority students (\autoref{sec:urm}).

@@

You should talk about things that adcom members might not know about and can help make you \textbf{stand out} in the application pool of thousands of applicants, e.g., you are member of a minority or LGBTQ+ group (\autoref{sec:urm}), your personal Github project with hundreds of stars or your regular contributions to well-known open-source projects (see \autoref{sec:improve-your-chance} for increasing your admission chance).

@@

Many universities now require a diversity, equity, and inclusion (DEI) statement as part of the application (also see \autoref{sec:urm}). Students, especially international ones, often struggle with what to write, especially when they are not from a minority group or do not have much experience in DEI.

@@

There are other things you might not think are important but can make you stand out. For example, if you have a strong background in a non-CS field but can be integrated with CS, e.g., you have a degree or background in \emph{dance} or \emph{music} and want to integrate them with CS, do talk about it. Are you a female or a minority in CS? Do you participate in outreach activities that help increase diversity and inclusion in CS? Diversity is valued in CS programs in the US (\autoref{sec:urm}), and experiences in these areas can make you stand out and get noticed from reviewers.

@@

For external fellowships including those from the US government (e.g., NSF GRFP, DOD NDSEG) and tech companies (e.g., Google, Microsoft, Facebook), you will need to apply.  % However, tech companies including Google, Microsoft, Facebook have fellowships that international students can apply for.
Prestigious external fellowships typically require a clear and good research plan (the GRFP also requires broader impacts discussion). So it might be a good idea to wait until your second year to have research experience and even publication before applying. Remember, you're competing with the top PhD students at top universities worldwide.

@@

In addition to general fellowships that are available to all students, there are also specific fellowships, e.g., for US citizens and permanent residents (\autoref{chap:domestic-students}), for underrepresented minority (URM) students (\autoref{sec:urm}), and students from specific countries. There are many fellowships available, so you should look for them and apply.

@@

I skim weak \emph{SOPs} but read strong ones carefully. I note whether the SOP is exciting, research-oriented, standing out, e.g., applicant is minority (\autoref{sec:urm}), and tailored to our program (e.g., if they are familiar with the work of some faculty or have talked to them). These notes are entered into the evaluation system.

@@

\section{Member of an underrepresented or LGBTQ+ groups}\label{sec:urm}
\sectioninfo{Being a minority can make you stand out and help your application.}

A common question is whether you should mention, e.g., in your SOP, that you are a member of an underrepresented and minority (URM) group\footnote{URM in CS in the US include African Americans, Hispanic Americans, American Indians/Native Alaskans, Native Hawaiians/Pacific Islanders (excluding Asian Americans), and multiracial students identifying with at least one of the previously listed URM categories.} or LGBTQ+ group.
The main concern is that you can be discriminated against or that you are asking for special treatment.

%This is a strong concern for international (and even domestic) students, who may have encountered discrimination in their countries or personal experiences. 
%This question arises often for international (and even domestic) students. The main worry is that they might get discriminated against, as often happen in their countries or personal experiences. %, and (ii) mentioning this might make them look like they are asking for special treatment.

In my personal opinion, highlighting URM or LGBTQ+ identity can \emph{be beneficial}, especially if you can articulate how your diverse experiences contribute to diversity and inclusion in academics.  Many universities have a strong commitment to diversity and inclusion and actively
\emph{recruit students from underepresentative backgrounds} (e.g., some \href{https://cse-climate.engin.umich.edu/reports/climate-dei-reports/cse-climate-and-dei-report-2022-2023/#grad-ethnicity}{stats from UMich CSE} and GMU often touts itself as one of the most diverse universities in the US). Moreover, many scholarships and fellowships are created specifically for URM and LGBTQ+ students, which you should explore if qualify.

Understandably, you might feel uncomfortable disclosing this, fearing you would lose a chance of working with a faculty due to their bias.  However, if someone has this issue, then you might not want to work with them anyway (and such bias is not acceptable in academics and likely would not be tolerated by the university). In my experience, it is often the opposite: \emph{many profs. actively seek to work with students from diverse backgrounds and view diversity as an asset to their research group}.

Moreover, promoting diversity aligns with the goal of broadening participation in computing (BPC) that is strongly supported within academia (e.g., see the \href{https://plans.bpcnet.org/GeorgeMasonUniversity_ComputerScience_DepartmentalBPCPlan.pdf}{BPC plan at Mason CS}) and government funding agencies such as NSF (e.g., see \href{https://new.nsf.gov/cise/broadening-participation}{here} and \href{https://www.nsf.gov/pubs/2022/nsf22125/nsf22125.jsp}{here}).
Providing this information can also make it easier for faculty to apply for certain scholarships or fellowships to support you, e.g., from a private donor that has offered a fellowship only to students from certain backgrounds.

Ultimately, this is a personal choice and you do what you feel comfortable.
If you do not feel comfortable disclosing this, you should not feel pressured to do so.
However, if you feel comfortable but are unsure if it is accepted in higher education in the US, then I hope this helps you understand that it is not only accepted but also valued and appreciated.

@@

  \item [Diversity Statement~\autoref{sec:diversity-statement}] A document sometimes required in applications where applicants describe their contributions to diversity and inclusion, or how their background adds to the diversity of the academic community. This document is often required for faculty positions, but more universities are requiring it for graduate admissions. Note that you do not need to be from an underrepresented group to write a diversity statement.

@@

Most of what is written in this handbook applies to both domestic\footnote{As mentioned in \autoref{chap:glossary}, domestic means you do not require a visa to study in the US.} and international students.  However, there are some differences and benefits that domestic students should be aware of and can leverage to improve their chances of admission.
\paragraph{Standing out \autoref{sec:improve-your-chance}} There are \emph{few} domestic applications compared to international ones, i.e., domestic students are the minority (\autoref{sec:urm}) in the CS PhD application pool. Many US universities want to increase the number of domestic students in their programs (and as mentioned later, there are specific fellowships and funding for domestic students).
That makes you stand out and can help your case.

>>>>

International students, especially those from less well-known universities and smaller countries, often wonder if they should apply to a CS PhD program in the US. Common concerns include (i) the \textbf{difficulty of getting admitted} and (ii) the \textbf{cost of graduate study} in the US. %, and for some, (iii) the fear of \textbf{discrimination} e.g., based on gender identity or cultural background.

@@

% \paragraph{Discrimination is not tolerated} Not only is discrimination illegal, diversity and inclusion are highly valued in US universities. Many universities take great pride in their diverse student body and faculty.

% In fact, many resources are available to encourage and support students from minority and diverse groups to pursue higher education and research, e.g., specific scholarships and fellowships. Faculty in CS are also familiar with and are encouraged to support such students in their group. There are also incentives, such as dedicated funding and awards, to motivate faculty to recruit and mentor minority students (\autoref{sec:urm}).

@@

%you are member of a minority or LGBTQ+ group (\autoref{sec:urm}),

You should talk about things that adcom members might not know about and can help make you \textbf{stand out} in the application pool of thousands of applicants, e.g., your personal Github project with hundreds of stars or your regular contributions to well-known open-source projects (see \autoref{sec:improve-your-chance} for increasing your admission chance).

@@

Many universities now require a diversity, equity, and inclusion (DEI) statement as part of the application. Students, especially international ones, often struggle with what to write, especially when they are not from a minority group or do not have much experience in DEI.

@@

There are other things you might not think are important but can make you stand out. For example, if you have a strong background in a non-CS field but can be integrated with CS, e.g., you have a degree or background in \emph{dance} or \emph{music} and want to integrate them with CS, do talk about it. Are you a female or a minority in CS? Do you participate in outreach activities that help increase diversity and inclusion in CS? Diversity is valued in CS programs in the US, and experiences in these areas can make you stand out and get noticed from reviewers.

@@

For external fellowships including those from the US government, e.g., see~\autoref{chap:fellowships} for NSF GRFP and  DOD NDSEG, and tech companies (e.g., Google, Microsoft, Facebook), you will need to apply.  % However, tech companies including Google, Microsoft, Facebook have fellowships that international students can apply for.
Prestigious external fellowships typically require a clear and good research plan (the GRFP also requires broader impacts discussion). So it might be a good idea to wait until your second year to have research experience and even publication before applying. Remember, you're competing with the top PhD students at top universities worldwide.

@@

%In addition to general fellowships that are available to all students, there are also specific fellowships, e.g., for US citizens and permanent residents (\autoref{chap:domestic-students}), for underrepresented minority (URM) students (\autoref{sec:urm}), and students from specific countries. There are many fellowships available, so you should look for them and apply.

@@

I skim weak \emph{SOPs} but read strong ones carefully. I note whether the SOP is exciting, research-oriented, standing out, and tailored to our program (e.g., if they are familiar with the work of some faculty or have talked to them). These notes are entered into the evaluation system.

@@

% \section{Member of an underrepresented or LGBTQ+ groups}\label{sec:urm}
% \sectioninfo{Being a minority can make you stand out and help your application.}

% A common question is whether you should mention, e.g., in your SOP, that you are a member of an underrepresented and minority (URM) group\footnote{URM in CS in the US include African Americans, Hispanic Americans, American Indians/Native Alaskans, Native Hawaiians/Pacific Islanders (excluding Asian Americans), and multiracial students identifying with at least one of the previously listed URM categories.} or LGBTQ+ group.
% The main concern is that you can be discriminated against or that you are asking for special treatment.

% %This is a strong concern for international (and even domestic) students, who may have encountered discrimination in their countries or personal experiences. 
% %This question arises often for international (and even domestic) students. The main worry is that they might get discriminated against, as often happen in their countries or personal experiences. %, and (ii) mentioning this might make them look like they are asking for special treatment.

% In my personal opinion, highlighting URM or LGBTQ+ identity can \emph{be beneficial}, especially if you can articulate how your diverse experiences contribute to diversity and inclusion in academics.  Many universities have a strong commitment to diversity and inclusion and actively
% \emph{recruit students from underepresentative backgrounds} (e.g., some \href{https://cse-climate.engin.umich.edu/reports/climate-dei-reports/cse-climate-and-dei-report-2022-2023/#grad-ethnicity}{stats from UMich CSE} and GMU often touts itself as one of the most diverse universities in the US). Moreover, many scholarships and fellowships are created specifically for URM and LGBTQ+ students, which you should explore if qualify.

% Understandably, you might feel uncomfortable disclosing this, fearing you would lose a chance of working with a faculty due to their bias.  However, if someone has this issue, then you might not want to work with them anyway (and such bias is not acceptable in academics and likely would not be tolerated by the university). In my experience, it is often the opposite: \emph{many profs. actively seek to work with students from diverse backgrounds and view diversity as an asset to their research group}.

% Moreover, promoting diversity aligns with the goal of broadening participation in computing (BPC) that is strongly supported within academia (e.g., see the \href{https://plans.bpcnet.org/GeorgeMasonUniversity_ComputerScience_DepartmentalBPCPlan.pdf}{BPC plan at Mason CS}) and government funding agencies such as NSF (e.g., see \href{https://new.nsf.gov/cise/broadening-participation}{here} and \href{https://www.nsf.gov/pubs/2022/nsf22125/nsf22125.jsp}{here}).
% Providing this information can also make it easier for faculty to apply for certain scholarships or fellowships to support you, e.g., from a private donor that has offered a fellowship only to students from certain backgrounds.

% Ultimately, this is a personal choice and you do what you feel comfortable.
% If you do not feel comfortable disclosing this, you should not feel pressured to do so.
% However, if you feel comfortable but are unsure if it is accepted in higher education in the US, then I hope this helps you understand that it is not only accepted but also valued and appreciated.

@@

  \item [Diversity Statement~\autoref{sec:diversity-statement}] A document sometimes required in applications where applicants describe their contributions to diversity and inclusion, or how their background adds to the diversity of the academic community. This document is often required for faculty positions, but some universities are requiring it for graduate admissions. 
  Note that you do not need to be from an underrepresented group to write a diversity statement.

@@

Most of what is written in this handbook applies to both domestic\footnote{As mentioned in \autoref{chap:glossary}, domestic means you do not require a visa to study in the US.} and international students.  However, there are some differences and benefits that domestic students should be aware of and can leverage to improve their chances of admission.
\paragraph{Standing out \autoref{sec:improve-your-chance}} There are \emph{few} domestic applications compared to international ones, i.e., domestic students are the minority in the CS PhD application pool. Many US universities want to increase the number of domestic students in their programs (and as mentioned later, there are specific fellowships and funding for domestic students).
That makes you stand out and can help your case.
